\documentclass[11pt,spaceforname]{../../jh/examjh}
\usepackage{../../../linalgjh}
\examhead{2007.1.Test1}{Questions}
\begin{document}
\begin{questions}
\question
Determine the coefficients $c_1$, $c_2$, and~$c_3$ for the cubic
polynomial~$p$ defined by the equation $p(x)=c_1x+c_2x^2+c_3x^3$
if the graph of~$p$ passes through the points $(-1,4)$, $(1,0)$,
and~$2,6$.
Proceed by obtaining a linear system of equations for the coefficients,
reformulating this system as a single matrix equation, and solving
the matrix equation by row-reduction.


\question
A linear system of equations is equivalent to the single matrix equation
$A\vec{x}=\vec{b}$.
Suppose that
\begin{equation*} 
  [A|\vec{b}]_\text{\textit{rref}}=
  \begin{amat}{5}
    1 &2 &0 &0 &3 &6 \\
    0 &0 &1 &0 &4 &7 \\
    0 &0 &0 &1 &5 &8 \\
    0 &0 &0 &0 &0 &0
  \end{amat}.
\end{equation*}
  \begin{parts}
    \item Determine all solutions of the linear system in vector form.
    \item Find all solutions of the equation $A\vec{x}=\vec{0}$.
  \end{parts}


\question
Suppose that $\map{f}{\Re^2}{\Re^2}$ is linear,
$f(\colvec{1 \\ 2})=\colvec{3 \\ 1}$, 
and $f(\colvec{1 \\ 4})=\colvec{1 \\ 3}$.
Determine the $\nbyn{2}$ matrix~$A$ corresponding to $f$ in two 
different ways.
\begin{parts}
  \item Use linearity to find $f(\colvec{1 \\ 0})$ and $f(\colvec{0 \\ 1})$
    and from this find~$A$.
  \item Convert the pair of given vector equations to a single matrix 
    equation for $A$ and solve it.
\end{parts}



\question
The equation $2x+y=0$ describes a line $L$ through the origin in $\Re^2$.
Find the matrix that corresponds to reflection across~$L$ and use it to
find the relected image of an arbitrary point $(r,s)$.



\question
Explain each of the following assertions.
\begin{parts}
\item Suppose that $\vec{u}$, $\vec{v}$, and~$\vec{w}$ are three vectors
in $\Re^2$ and no pair of them is collinear.
There are always infinitely many different linear combinations of these
three vectors that sum to the zero vector.

\item Projection onto a line in $\Re^3$ is not an invertible 
transformation.  
\end{parts}
\end{questions}
\end{document}
